Open Clusters Characterization in Gaia DR2 Using ML Algorithms

### 2020-11-29 thesis/thesis.tex
thesis: Update Method chapter

The first one shows a resonance at $\approx 7.5mas$ belonging the OC. While the second figure would help us to look for isochrone curves and so,
to determine the age of these stars.

@@

\section{Feature Selection}

@@

So, as summary, these are the features we are going to use as sources for our clustering models:

\begin{itemize}
  \item \emph{Proper motion in right ascension}: $\mu_{\alpha}$
  \item \emph{Proper motion in declination}: $\mu_{\delta}$
  \item \emph{Parallax}: $\varpi$
\end{itemize}

This feature selection has been refined by iterating over the K-Means clustering process with a fixed number of clusters and varying the feature selection in each case.
The final selection is the one with best silhouette score. This process is explained in more detail in Appendix \ref{chap:feature_selection}.

@@

Once we have found the set of features that best describes our problem, we can begin to search for the OC in the study region.

@@

Since we are looking for a single cluster, it seems reasonable to use a clustering algorithm and set it to find two clusters.
One for the desired OC and another cluster which contains stars that don't belong to the open cluster.
However, this idea is not completely right.

@@

As shown in Figure \ref{fig:kmeans_comparisons_melotte_22}, bigger values for the number of clusters allow us to isolate more accurately
the resonance in parallax at $\approx 7.5mas$. However, we have the disadvantage that more groups are formed, while our aim is to find just
one small cluster and another with the rest of stars. This effect complicates our task of finding the desired OC. Therefore, we have to find
a way to set the right value for the number of clusters to isolate the searched cluster without creating to many groups.

To solve this issue, we will try to estimate the best number of clusters by using the \emph{silhouette score}.

\section{Deep Embedded Clustering}

Our initial approach allows us to find some groups that potentially contain the desired OC. However we need now a method for merging clusters
by migrating stars from one group to another. So we can end up with the minimum number of clusters while preserving the open cluster.

Since we don't have a labeled dataset which allows as to train a supervised model, we are required to use an unsupervised self-trained model.

For that purpose we have adapted the Unsupervised Deep Embedding for Clustering Analysis or DEC model \cite{xie2016unsupervised} to our work.

@@

Although in this work we are managing a small number of features, this latent space allows us to reduce the number of features, avoiding
this way the \emph{"curse of dimensionality"} \cite{bellman1961curse}.

These encoders are pretrained before fitting the model to generate predictions. Then, the encoder is used to transform input data to the
latent space $Z$. Once the data has been transformed, a K-Means clusterer is used to make an initial clustering of the data.

@@

In the soft assignment stage, the \emph{Student's t-distribution} is used as a kernel to measure the similarity between the embedded points
and the cluster centroid. While in the KL divergence minimization the algorithm iteratively refines clusters by learning from their high
confidence assignments with the help of an auxiliary target distribution. The model is trained by matching the soft assignment to the target
distribution. The choice of this target distribution is crucial for DEC's performance. In this work we have taken the target distribution
from DEC's original paper \cite{xie2016unsupervised}, which is defined as follows:

\begin{equation}
  p_{ij} = \frac{q^{2}_{ij} / f_{j}}{\sum_{j'}q^{2}_{ij'}/f_{j'}}
\end{equation}

Inside \href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/notebooks}{src/notebooks},
there is a notebook called
\href{https://github.com/cdalvaro/machine-learning-master-thesis/blob/main/src/notebooks/cluster_characterization.ipynb}{cluster\_characterization.ipynb}
that illustrates how DEC model processes a region of stars trying to characterize the OC hidden inside that region.

As we can see in Figure \ref{fig:dec_melotte_22}, the DEC model is able to merge similar clusters leaving alone the OC. Which is exactly what we are looking for.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_pm_melotte_22.pdf}
      \caption{Proper motions}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_parallax_melotte_22.pdf}
      \caption{Parallax}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_isochrone_melotte_22.pdf}
      \caption{Isochrone curve}
    \end{subfigure}
  \end{subfigure}
  \caption{DEC model applied to Melotte 22}
  \label{fig:dec_melotte_22}
\end{figure}

>>>>

The figure on the left shows a resonance at $\approx 7.5mas$ belonging the OC. While the figure on the right would help us to look for isochrone curves and so,
to determine the age of these stars.

@@

\section{Feature Selection}
\label{sec:feature_selection}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/features_melotte_22.pdf}
  \end{subfigure}
  \caption{Feature pairplot}
  \label{fig:features_melotte_22}
\end{figure}

So, in summary, these are the features we are going to use as sources for our clustering models:

\begin{itemize}
  \item \emph{Proper motion in right ascension} (corrected by parallax): $\mu_{\alpha}$
  \item \emph{Proper motion in declination} (corrected by parallax): $\mu_{\delta}$
  \item \emph{Parallax}: $\varpi$
  \item \emph{Proper motion modulus}: $\left\| \vec{\mu} \right\|$
\end{itemize}

This feature selection has been refined by iterating over the K-Means clustering process with a fixed number of clusters and varying the feature selection in each case.
The final selection is the one with best \emph{silhouette score} \cite{rousseeuw1987silhouettes}.

This process is explained in more detail in Appendix \ref{chap:feature_selection}.

@@

Once we have found the set of features that best describes our problem, we can begin searching for the OC in the study region.

@@

Since we are looking for a single cluster, it seems reasonable to use a clustering algorithm and set it to find two clusters.
One for the desired OC and another which contains stars that do not belong to the open cluster.
However, this idea is not completely right.

@@

As shown in Figure \ref{fig:kmeans_comparisons_melotte_22}, bigger values for the number of clusters allow us to isolate more accurately
the resonance in parallax at $\approx 7.5mas$. However, we have the disadvantage that more groups are formed, while our aim is to find just
one small cluster that contains the stars that belong to the OC and another one with the remaining stars.

This effect complicates our task of finding the desired OC. Therefore, we have to find
a way to set the right value for the number of clusters to isolate the searched cluster without creating too many groups.

To solve this issue, we will try to estimate the best number of clusters by using one more time the \emph{silhouette score}.

The following example can be found in the Jupyter notebook: \verb|cluster_characterization.ipynb| (available at \verb|src/notebooks|).
It shows how to estimate the best number of clusters for Melotte 22 by using \verb|cdalvaro.ml.estimate_n_clusters| method.

\begin{minted}{python}
  n_clusters, kmeans = estimate_n_clusters(melotte22_df, min_clusters=3,
                                           max_clusters=10, verbose=True)

  # Output
  # Silhouette score for 3 clusters: 0.5420
  # Silhouette score for 4 clusters: 0.5393
  # Silhouette score for 5 clusters: 0.5608
  # Silhouette score for 6 clusters: 0.5336
  # Silhouette score for 7 clusters: 0.5306
  # Silhouette score for 8 clusters: 0.4978
  # Silhouette score for 9 clusters: 0.5214
  # Silhouette score for 10 clusters: 0.5199
  # Best silhouette score is 0.5608 for 5 clusters
\end{minted}

K-Means does a good job making an initial clustering, as shown in Figure \ref{fig:kmeans_melotte_22}. However, too many clusters arise from this characterization
and the open cluster is still polluted with stars that do not belong to it. Even more, we would like to reduce the amount of clusters too.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_pm_melotte_22.pdf}
      \caption{Proper Motion}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans_parallax_melotte_22.pdf}
      \caption{Parallax}
    \end{subfigure}
  \end{subfigure}
  \caption{K-Means model applied to Melotte 22}
  \label{fig:kmeans_melotte_22}
\end{figure}

If we take a look into the H-R diagram (Figure \ref{fig:kmeans_hr_diagram_melotte_22}), we can identify the isochrone curve for Melotte 22 cluster.
This curve has a good shape, but again, it contains outsider stars. Therefore, we would like to find a better model that improves this initial characterization
by reducing the amount of clusters and also that would be able to remove outsiders from the OC.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/kmeans_isochrone_melotte_22.pdf}
  \end{subfigure}
  \caption{Melotte 22 H-R diagram with K-Means characterization}
  \label{fig:kmeans_hr_diagram_melotte_22}
\end{figure}

\section{Deep Embedded Clustering (DEC)}

Our initial approach allows us to find some groups that potentially contain the desired OC. However, as we have seen, the result is not enough
accurate. Therefore, we need a method for merging clusters by migrating stars from one group to another,
so we can end up with the minimum number of clusters while preserving the open cluster.

Since we do not have a labeled dataset which allows us to train a supervised model, we have no choice but to use an unsupervised self-trained model.

For that reason we have adapted the \emph{Unsupervised Deep Embedding for Clustering Analysis} or \emph{DEC} model \cite{xie2016unsupervised} to our work.

@@

Although, as explained in section \ref{sec:feature_selection}, the number of features we are managing is not too large, this latent space allows us to reduce
the number of features, avoiding this way the \emph{"curse of dimensionality"} \cite{bellman1961curse}.

These encoders are pretrained before fitting the model to generate predictions. Then, the encoder is used to transform input data to the
latent space $Z$. Once the data has been transformed, a K-Means clusterer is used to make an initial clustering.

@@

\begin{equation}
  p_{ij} = \frac{q^{2}_{ij} / f_{j}}{\sum_{j'}q^{2}_{ij'}/f_{j'}}
  \label{eq:student_tdistribution}
\end{equation}

In the soft assignment stage, the \emph{Student's t-distribution} is used as a kernel to measure the similarity
between the embedded points and the cluster centroid. While in the KL divergence minimization the algorithm iteratively refines clusters by learning
from their high confidence assignments with the help of an auxiliary target distribution. The model is trained by matching the soft assignment to the
target distribution. The choice of this target distribution is crucial for DEC's performance. In this work we have taken the target distribution
from DEC's original paper \cite{xie2016unsupervised}, which is defined in Equation \ref{eq:student_tdistribution}.

Figure \ref{fig:dec_model_setup} shows the layer setup of our DEC model. It is simpler than the one tested on the original paper \cite{xie2016unsupervised},
since the number of selected features in our work is smaller. Therefore, using the same configuration would result in a model too
powerful that would incur in overfitting issues unable to make right predictions.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \includegraphics[width=\textwidth]{../figures/dec_diagram.pdf}
  \end{subfigure}
  \caption{DEC model layer setup}
  \label{fig:dec_model_setup}
\end{figure}

Once the model is ready, we can test it by applying it to the Melotte 22 dataset. Notebook \verb|cluster_characterization.ipynb| illustrates how DEC model
processes a region of stars trying to characterize the open cluster hidden within that region.

As shown in Figure \ref{fig:dec_melotte_22}, the DEC model is able to move stars from one group to another removing some clusters and improving the OC characterization.
This is exactly what we are looking for.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.9\textwidth}
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_pm_melotte_22.pdf}
      \caption{Proper Motion}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_parallax_melotte_22.pdf}
      \caption{Parallax}
    \end{subfigure}
  \end{subfigure}
  \caption{DEC model applied to Melotte 22}
  \label{fig:dec_melotte_22}
\end{figure}

The H-R diagram has been improved as well, and the isochrone curve of the open cluster is sharper than the one obtained with the K-Means algorithm
(Figure \ref{fig:kmeans_hr_diagram_melotte_22}).

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/dec_isochrone_melotte_22.pdf}
  \end{subfigure}
  \caption{Melotte 22 H-R diagram with DEC characterization}
  \label{fig:dec_hr_diagram_melotte_22}
\end{figure}
